---
title: "Assignment1"
output: html_document
---

Obtained activity data from: [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 

Data was logged in 5 minute intervals.

```{r}

data <- read.csv("activity.csv")
nrow(data)

steps <- data$steps
sum(steps, na.rm = TRUE)
days <- split(steps, data$date)
numDays <- length(days)
numDays

rowsPerDay <- nrow(data) / length(days)
rowsPerDay

```

### Means and Medians

```{r}

# Make a histogram of the total number of steps taken each day

# sums
daySums <- sapply(1:length(days), function (d) sum (days[[d]], rm.na = TRUE))
hist(daySums)

# Calculate and report the mean and median total number of steps taken per day

# means
dayMeans <- sapply(1:length(days), function (d) mean (days[[d]], rm.na = TRUE))
dayMeans

# medians
dayMedians <- sapply(1:length(days), function (d) median (days[[d]], na.rm = TRUE))
dayMedians

```

### Mean per 5 minute interval

```{r}

startIndices <- (0 : (numDays - 1)) * rowsPerDay
intervalsFunc <- function (i) startIndices + i
intervals <- sapply(1:rowsPerDay, intervalsFunc)

f <- function (x) {
  ns <- steps[intervals[,x]] # get interval across all days
  ns <- ns[!is.na(ns)] # remove nas
  if(length(ns) > 0){
    mean(ns)
  } else {
    0 # if there is no data for this day, return a default of 0
  }
}

# the mean for each of the intervals
imeans <- sapply(1:rowsPerDay, f)

# Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

graphics::plot.default(
  x = 1:rowsPerDay,
  y = imeans,
  type = "l",
  xlab = "5 minute interval",
  ylab = "Mean",
  main = "Mean per 5 minute interval"
)

# Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

max(imeans) # the highest average number of steps
which.max(imeans) # this interval had the highest average number of steps

```

### Missing Values

```{r}

# Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

# total number of missing values in the dataset
length(steps[is.na(steps)])

# converts an index into one of the intervals
index2interval <- function (i) {
  day <- ceiling(i / rowsPerDay)
  i - ((day - 1) * rowsPerDay)
}

# converts an index into a value. 
# if the step data exists, it will be returned, 
# otherwise the average for that interval across all days is returned
f <- function (x) {
  entry <- steps[[x]]
  if(is.na(entry)){
    q <- index2interval(x);
    z <- imeans[q]
    z
  } else {
    entry
  }
}

# Devise a strategy for filling in all of the missing values in the dataset.
# Create a new dataset that is equal to the original dataset but with the missing data filled in.

# creates a new vector of step data where missing input is replaced by the average at that interval across all days 
newSteps <- sapply(1:length(steps), f)
sum(newSteps)
newDays <- split(steps, data$date)

# Make a histogram of the total number of steps taken each day 
newDaySums <- sapply(1:length(newDays), function (d) sum (newDays[[d]]))
hist(newDaySums)

# Calculate and report the mean and median total number of steps taken per day. 

# means
newDayMeans <- sapply(1:length(newDays), function (d) mean (newDays[[d]]))
newDayMeans

# medians
newDayMedians <- sapply(1:length(newDays), function (d) median (newDays[[d]]))
newDayMedians

```

*Do these values differ from the estimates from the first part of the assignment?*

The new means and medians appear to be similar to the original data.

*What is the impact of inputing missing data on the estimates of the total daily number of steps?*

Inputing missing data did not have a big impact on the total daily number of steps.

### Are there differences in activity patterns between weekdays and weekends?

Data starts on Monday (Oct 1, 2012) and ends Friday (Nov 30, 2012).

The data spans 8 weeks and 5 weekdays, so there should be 45 weekdays and 16 weekends.

```{r}

f <- function (x) {
  
  date <- as.Date(data$date[x])
  dayName <- weekdays(date)
  if(dayName == "Saturday" || dayName == "Sunday"){
    "weekend"
  } else {
    "weekday"
  }
   
}

# Create a new factor variable in the dataset with two levels - "weekday" and "weekend"

dayTypes <- sapply(1 : nrow(data), f)

length(dayTypes[dayTypes == "weekday"]) / rowsPerDay
length(dayTypes[dayTypes == "weekend"]) / rowsPerDay

```

### Mean per 5 minute interval for weekdays 

```{r}

newData <- cbind(newSteps, dayTypes)
newData <- as.matrix(newData)

# Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

weekDaySteps <- newData[newData[,2] == "weekday", 1]
newNumDays <- length(weekDaySteps) / rowsPerDay
newNumDays

startIndices <- (0 : (newNumDays - 1)) * rowsPerDay
intervalsFunc <- function (i) startIndices + i
intervals <- sapply(1:rowsPerDay, intervalsFunc)

f <- function (x) {
  ns <- as.numeric(weekDaySteps[intervals[,x]]) # get interval across all days
  ns <- ns[!is.na(ns)] # remove nas
  if(length(ns) > 0){
    mean(ns)
  } else {
    0 # if there is no data for this day, return a default of 0
  }
}

# the mean for each of the intervals
imeans <- sapply(1:rowsPerDay, f)

# Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

graphics::plot.default(
  x = 1:rowsPerDay,
  y = imeans,
  type = "l",
  xlab = "5 minute interval",
  ylab = "Mean",
  main = "Mean per 5 minute interval for weekdays"
)


```

### Mean per 5 minute interval for weekends

```{r}

weekEndSteps <- newData[newData[,2] == "weekend", 1]
newNumDays <- length(weekEndSteps) / rowsPerDay
newNumDays

startIndices <- (0 : (newNumDays - 1)) * rowsPerDay
intervalsFunc <- function (i) startIndices + i
intervals <- sapply(1:rowsPerDay, intervalsFunc)

f <- function (x) {
  ns <- as.numeric(weekEndSteps[intervals[,x]]) # get interval across all days
  ns <- ns[!is.na(ns)] # remove nas
  if(length(ns) > 0){
    mean(ns)
  } else {
    0 # if there is no data for this day, return a default of 0
  }
}

# the mean for each of the intervals
imeans <- sapply(1:rowsPerDay, f)

# Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

graphics::plot.default(
  x = 1:rowsPerDay,
  y = imeans,
  type = "l",
  xlab = "5 minute interval",
  ylab = "Mean",
  main = "Mean per 5 minute interval for weekends"
)


```

### End.